A Formalisation of Sturm's theorem in \isabellehol

### 2013-12-05 paper/paper.tex
Added content to slides, tuned paper and guide

\begin{lemma}\label{thm:sturm_multiple_aux}
Let $P\in\RR[X]\setminus\RR$, i.e. a nonconstant real polynomial. Let $P_0,\ldots,P_{n-1}$ be the result of the canonical Sturm sequence construction of $P$ and define $Q_i := P_i / \gcd(P,P')$. Then $Q_i$ is a Sturm sequence and $Q_0$ has the same roots, disregarding multiplicity, as $P$.
\end{lemma}
\begin{proof}
First, we note that $\gcd(P,P')\neq 0$, since $P'\neq 0$ by assumption. Furthermore, lemma \ref{thm:sturm_gcd} implies that $\gcd(P,P')\ |\ P_i$ for all $i$, thus $Q_i\in\RR[X]$, i.e. the $Q_i$ really are polynomials. This means that our definition of $Q_0,\ldots,Q_{n-1}$ is well-defined.\\
Now we show that the five properties for a Sturm sequence are satisfied:
\begin{itemize}
\item the sequence has at least length 2: obvious by construction of $Q_0, \ldots, Q_{n-1}$
\item $Q_{n-1}$ does not change its sign: obvious, since $P_{n-1}$ is constant, so $Q_{n-1}$ is, too.
\item $Q_0$ and $Q_1$ have no common roots:\\
By construction, we have $Q_0 = P / \gcd(P,P')$ and $Q_1 = (P / \gcd(P, P'))'$. This then implies $\gcd(Q_0, Q_1) = 1$. However, the proof for this algebraic fact is nontrivial; it was, of course, proven in the \emph{Isabelle} version of this proof, but it is too complex and at the same time not important enough to be discussed in detail here.
\item for any $i\in\{0,\ldots,n-1\}$ and any root $x_0$ of $Q_{i+1}$, $Q_{i}(x_0)Q_{i+2}(x_0) < 0$ holds:\\
By construction of the canonical Sturm sequence, we have:
$$P_{i+2} = -P_{i}\ \mathrm{mod}\ P_{i+1}$$
Therefore, we have:
$$(P_{i}\ \mathrm{div}\ P_{i+1}) \cdot P_{i+1} - P_{i+2} = P_{i}$$
Let $D:=\gcd(P,P')$. Since $D$ divides all the $P_j$, we have $P_{i}=D\cdot Q_{i}$, $P_{i+1}=D\cdot Q_{i+1}$, and $P_{i+2}=D\cdot Q_{i+2}$ and thus:
$$(D\cdot Q_{i}\ \mathrm{div}\ D\cdot Q_{i+1}) \cdot D\cdot Q_{i+1} - D\cdot Q_{i+2} = D\cdot Q_{i}$$
Cancelling $D$ (allowed since $\RR[X]$ is an integral domain and $D\neq 0$) yields:
$$(Q_{i}\ \mathrm{div}\ Q_{i+1}) \cdot Q_{i+1} - Q_{i+2} = Q_{i}$$
Since $x_0$ is a root of $Q_{i+1}(x_0)$, we then have:
$$-Q_{i+2}(x_0) = Q_{i}(x_0)$$
Since $\gcd(Q_{i}, Q_{i+1}) = 1$ and $x_0$ is a root of $Q_{i+1}(x_0)$, $x_0$ cannot be a root of $Q_{i}$ and thus we have:
$$Q_{i}(x_0)Q_{i+2}(x_0) = -Q_{i}(x_0)^2 < 0$$

>>>>

\begin{lemma}\label{thm:sturm_multiple_aux}
Let $P\in\RR[X]\setminus\RR$, i.e. a nonconstant real polynomial. Let $P_0,\ldots,P_{n-1}$ be the result of the canonical Sturm sequence construction of $P$ and define $Q_i := P_i / \gcd(P,P')$. Then $Q_i$ is a Sturm sequence and $Q_0$ has the same roots, disregarding multiplicity, as $P$.
\end{lemma}
\begin{proof}
First, we note that $\gcd(P,P')\neq 0$, since $P'\neq 0$ by assumption. Furthermore, lemma \ref{thm:sturm_gcd} implies that $\gcd(P,P')\ |\ P_i$ for all $i$, thus $Q_i\in\RR[X]$, i.e. the $Q_i$ really are polynomials. This means that our definition of $Q_0,\ldots,Q_{n-1}$ is well-defined.\\
Now we show that the five properties for a Sturm sequence are satisfied:
\begin{itemize}
\item the sequence has at least length 2: obvious by construction of $Q_0, \ldots, Q_{n-1}$
\item $Q_{n-1}$ does not change its sign: obvious, since $P_{n-1}$ is constant, so $Q_{n-1}$ is, too.
\item $Q_0$ and $Q_1$ have no common roots:\\
By construction, we have $Q_0 = P / \gcd(P,P')$ and $Q_1 = P' / \gcd(P, P')$. Obviously, $Q_0$ and $Q_1$ are then coprime and cannot have any common roots.
\item for any $i\in\{0,\ldots,n-1\}$ and any root $x_0$ of $Q_{i+1}$, $Q_{i}(x_0)Q_{i+2}(x_0) < 0$ holds:\\
By construction of the canonical Sturm sequence, we have:
$$P_{i+2} = -P_{i}\ \mathrm{mod}\ P_{i+1}$$
Therefore, we have:
$$(P_{i}\ \mathrm{div}\ P_{i+1}) \cdot P_{i+1} - P_{i+2} = P_{i}$$
Let $D:=\gcd(P,P')$. Since $D$ divides all the $P_j$, we have $P_{i}=D\cdot Q_{i}$, $P_{i+1}=D\cdot Q_{i+1}$, and $P_{i+2}=D\cdot Q_{i+2}$ and thus:
$$(D\cdot Q_{i}\ \mathrm{div}\ D\cdot Q_{i+1}) \cdot D\cdot Q_{i+1} - D\cdot Q_{i+2} = D\cdot Q_{i}$$
Cancelling $D$ (allowed since $\RR[X]$ is an integral domain and $D\neq 0$) yields:
$$(Q_{i}\ \mathrm{div}\ Q_{i+1}) \cdot Q_{i+1} - Q_{i+2} = Q_{i}$$
Since $x_0$ is a root of $Q_{i+1}(x_0)$, we then have:
$$-Q_{i+2}(x_0) = Q_{i}(x_0)$$
Since $\gcd(Q_{i}, Q_{i+1}) = 1$ and $x_0$ is a root of $Q_{i+1}(x_0)$, $x_0$ cannot be a root of $Q_{i}$ and thus we have:
$$Q_{i}(x_0)Q_{i+2}(x_0) = -Q_{i}(x_0)^2 < 0$$
